---
title: "Berkeley Madonna Conversion (first draft)"
author: "Mike K Smith"
date: "25 May 2018"
output:
  html_document:
    df_print: paged
---

## Getting BM Equations
Open the .mmd file in Berkeley Madonna and then from the "File" menu choose
"Save equations as..." and save the equations in .txt format.

## Read Berkeley Madonna file
*** Comments *** 
BM uses `;` or `{ ... }` as the comment delimiters. To facilitate conversion I have
omitted all comments from the text before processing. 

```{r readLines, message = FALSE, warning = FALSE}
library(tidyverse, quietly = TRUE)
txt <- readLines("Calcium Osc (phase Plane) Equations.txt") %>%
  str_replace("\\s","") %>%
  str_trim() %>%
  str_replace(";.*", "") %>%
  str_replace("\\{.*","") %>%
  .[. != ""]

head(txt)
```
### Convert ODEs to mrgsolve format
*** mrgsolve block ***  
The \$ODE block in mrgsolve contains the ODEs for the "compartments".

*** Assumptions ***  
LHS of the ODE contains the ODE compartment variable name.
RHS of the ODE can contain either other compartment variables / ODE amounts, 
model parameters and inputs / covariates.

*** Conversion detail ***  
BM ODEs in this text are of the form: `d/dt( <var> ) = ...`.
For mrgsolve they should be `dxdt_var = ...`

Pattern `\\(|\\)|\\*|\\/|\\-|\\+|\\>` should remove arithmetic relationships 
leaving only variable names and constants.

```{r mrgsolveODEs}
ODE_lines <- stringr::str_detect(txt, pattern="d/dt *\\( *")

ODEs <- txt[ODE_lines] %>% 
  stringr::str_replace(pattern = "d/dt *\\( *", replacement = "dxdt_") %>%
  stringr::str_replace(pattern = " *\\) *", replacement = "")

ODE_cmts <- stringr::word(ODEs,1,sep="=") %>%
  stringr::word(2,sep="_") %>%
  .[!is.na(.)]

print("$ODE")
ODEs
print("$CMT")
ODE_cmts

## Determine if character string is numeric or other
is_nonmissing_number <- function(x)!is.na(as.numeric(x))

## Take RHS of assignments and remove arithmetic operators
##   THEN reduce to the set of variables used in ODEs.
##   Eliminate any numeric constants

pattern <- "\\(|\\)|\\*|\\/|\\-|\\+|\\>|\\^"
ODE_vars <- ODEs %>%
  stringr::word(start=2, sep=" *= *") %>%
  stringr::str_replace_all(pattern = pattern, replacement = " ") %>%
  stringr::str_split(pattern = "[:space:]") %>%
  unlist() %>%
  unique() %>%
  .[.!=""] %>%
  .[!is_nonmissing_number(.)]

## Eliminate RHS variables that are compartments
ODE_vars <- ODE_vars[!(ODE_vars %in% ODE_cmts)]

print("Variables & parameters used in ODEs")
ODE_vars
```

### Find initial conditions
*** mrgsolve block ***  
For mrgsolve, if the initial condition = 0 then this compartment is initialized
in a \$CMT block. Otherwise it is initialized in a \$INIT block.

*** Assumptions ***  
BM has lines beginning with `init` to signify that these initialise ODE 
compartments. BM treats initialising with zero amount and initialising with
non-zero amount equivalently.

LHS of assignment should be an ODE variable (after the `init` statement).
RHS of assignment could be zero or could be a decimal amount.

*** Conversion details ***   
Assuming that lines begin `init <ODE_var>`. So ODE_var to be initialized is 
second "word".

Assuming only decimal values for non-zero initialized variables (rather than
assignment to a variable).

```{r initial_conditions}
## Initialisation lines start with "init"
init_lines <- casefold(txt, upper = FALSE) %>%
  stringr::str_detect(pattern = "^ *init")

if(sum(init_lines)>0){
  initialized <- txt[init_lines]
  
  initialized_vars <- stringr::word(initialized,1,sep="=") %>%
    stringr::word(2)
  
  # Take the RHS after the assignment and before the comment
  value <- initialized %>%
    stringr::word(2,sep="= *") %>%
    as.numeric()
  
  ## Only those initialisation lines where the value is non-zero
  ## These are initialised in $INITS block
  inits <- gsub("^ *init ", "", initialized[value != 0])
  
  print("$INITS")
  inits
}

## if the initialisation is to zero, then these are defined in $CMT block
cmts <- initialized[value==0]
if(length(cmts)>0){
  str_replace("^ *init ", "") %>%
  stringr::word(1,sep="=")

  print("$CMT")
  cmts
}

```

Next check if there are ODE compartments for which there is no value for initial 
condition.
```{r noInitialValue_ODE_vars}
uninitialized <- ODE_cmts[!is.element(ODE_cmts,initialized_vars)]
if(length(uninitialized) == 0) print("All compartments initialised")else(
    print(paste("The following compartments are uninitialized",uninitialized))
  )
```

### Tidy remaining assignments
There are many assignments that are fixed values, some expressed as fractions,
and some IF...THEN...ELSE statements.

*** Assumptions ***  
For IF...THEN...ELSE statements, then the "if" is the first "word" after the 
assignment statement i.e. "... = if ...".

*** Conversion details ***  
Look for assignment statements, and eliminate any that have IF...THEN...ELSE
structure.

Of those that remain, use eval(parse(<text>)) to attempt to calculate the RHS
of the assignment. Keep only those that work. $FIXED can take arithmetic 
assignment but not assignment to variables.

```{r tidyAssignments}
assignments <- txt[!(ODE_lines | init_lines)] %>%
  stringr::str_subset(pattern = " *= *")

assigned_vars <- assignments %>%
  stringr::word(start=1, sep=" *= *")

## Handle IF...THEN...ELSE statements
uses_if <- assignments %>%
  stringr::word(start=2, sep=" *= *") %>%
  stringr::str_detect(pattern="^ *if")

if_statements <- assignments[uses_if] %>%
  stringr::str_replace(" *if *","") %>%
  stringr::str_replace(" *then *"," ? ") %>%
  stringr::str_replace(" *else *"," : ")

if_vars <- assigned_vars[uses_if]

assignments2 <- assignments[!uses_if] 

RHS_works <- assignments2 %>%
  stringr::word(start=2, sep=" *= *") %>%
  map(~ try(eval(parse(text=.)),silent = TRUE)) %>%
  map( ~ !inherits(., "try-error")) %>%
  unlist()

fixed = assignments2[RHS_works]

fixed_vars <- assignments2[RHS_works] %>%
  stringr::word(.,start=1, sep=" *= *")  
```

### Separate \$MAIN, \$ODE and \$TABLE variables
***mrgsolve block***  
The mrgsolve \$MAIN block is analogous to the NONMEM \$PK block. This block 
defines variables that combine parameters, fixed values etc. Equations in this
block are evaluated just prior to advancing the ODEs in time.

The \$ODE block should contain variable which use compartment amounts, where 
those variables are themselves used in calculations of compartment amounts. 
Equations in this block are evaluated at each advancement of the ODEs.

The \$TABLE block can take variables which use ODE compartment amounts, or 
other variables, but these should not be used in definition or calculation of
ODE amounts or variables. Equations in the \$TABLE block are evaluated ***after***
the ODEs are advanced.

*** Asssumptions ***  
There are a few cases for \$MAIN variable definition:  
  1. functions of \$FIXED variables.
  2. functions of \$PARAM parameters.
  
For \$ODE variable definitions, they may use compartment amounts, but will also
define variables used within the system of ODEs.

\$TABLE variables may use any of the model variables, but these should not 
define any variables that are used in calculation of ODE variables or amounts.

Assignments which use ODEs compartment variables can either be:  
  * variables and values needed for \$ODE block specification OR
  * output variables which should be specified in the \$TABLE block.
  
If the variable uses ODE compartments in the RHS of the assignment AND the LHS
is NOT used in calculation of ODE amounts THEN we will assume this value is used
in \$TABLE.

*** Conversion details ***  
Look at RHS of assignments for \$FIXED variables or constants (numerics). These
should go into \$MAIN.

Determine if the RHS of the assignment contains an ODE compartment variable THEN
if the LHS is not used in calculation of ODEs (i.e. in ODE_vars set) AND is not
used in RHS assignment of ODE variables then push to \$TABLE.

```{r warning=FALSE, message=FALSE}
remainder <- assignments[!is.element(assigned_vars, fixed_vars)|uses_if]
remainder_vars <- remainder %>%
  stringr::word(start=1, sep=" *= *")

# subset uses_if to match remainder
uses_if <- uses_if[!is.element(assigned_vars, fixed_vars)|uses_if]

## Take RHS and find variables that need defined
pattern <- "\\(|\\)|\\*|\\/|\\-|\\+|\\>|\\?|\\:|\\^"
RHS1 <- remainder %>%
  stringr::word(start=2, sep=" *= *") %>%
  stringr::word(start=1, sep=" *#|;") %>%
  stringr::str_replace_all(pattern = pattern, replacement = " ") %>%
  stringr::str_replace_all(pattern = "\\s+", replacement = " ") %>%
  stringr::str_replace_all(pattern = " *if *", replacement = " ") %>%
  stringr::str_replace_all(pattern = " *then *", replacement = " ") %>%
  stringr::str_replace_all(pattern = " *else *", replacement = " ") %>%
  stringr::str_trim() %>%
  stringr::str_split(pattern = "[:space:]")
names(RHS1) <- remainder_vars

## Determine whether elements on RHS are functions of numbers and fixed variables 
is_simple <- sapply(RHS1,function(x){
  all(is_nonmissing_number(x) | x %in% fixed_vars)
})

main1 <- c(remainder[is_simple],if_statements)
main_vars1 <- c(remainder_vars[is_simple], if_vars)

remainder2 <- remainder[!(is_simple | uses_if)]
remainder2_vars <- remainder_vars[!(is_simple | uses_if)]
RHS2 <- RHS1[!(is_simple | uses_if)]

## Determine if remaining assignments use ODE compartment variables
uses_cmts <- RHS2 %>%
  stringr::str_detect(pattern = paste( ODE_cmts, collapse="|"))

uses_cmt_assignments <- remainder2[uses_cmts]
uses_cmt_vars <- remainder2_vars[uses_cmts]

## Determine if assignment defines variable used in ODEs
defines_ODE_variable <- is.element(uses_cmt_vars,ODE_vars)

ODE_assignments <- uses_cmt_assignments[defines_ODE_variable] %>%
  paste("double", . )
ODE_assignment_vars <- uses_cmt_vars[defines_ODE_variable]

ode <- c(ODEs, ODE_assignments)

## Determine if assignment defines variable used in 
##   definition of variable used in ODE
remainder3 <- remainder2[!defines_ODE_variable]
remainder3_vars <- remainder2_vars[!defines_ODE_variable]

used_by_ODE_var <- stringr::str_detect(RHS2[defines_ODE_variable],
                                       remainder3_vars)

main2 <- remainder3[used_by_ODE_var]
main2_vars <- remainder3_vars[used_by_ODE_var]
# 
# remainder3 <- remainder2[!uses_cmts]
# remainder3_vars <- remainder2_vars[!uses_cmts]
# RHS3 <- RHS2[!uses_cmts]
# 
# main2 <- remainder3[defines_ODE_var]
# main_vars2 <- remainder3_vars[defines_ODE_var]

## 
table1 <- uses_cmt_assignments[!defines_ODE_variable]
table_vars1 <- uses_cmt_vars[!defines_ODE_variable]

table2 <- remainder3[!defines_ODE_variable]
table_vars2 <- remainder3_vars[!defines_ODE_variable]
```
### \$ODE parameters to \$PARAM, all others to \$FIXED
Of the assigned variables, we can push any that are referenced in 
RHS of ODE equations to the \$PARAM block, and the remainder to \$FIXED. 

*** mrgsolve blocks ***  
mrgsolve has the \$PARAM block for any parameters used in the ODE equations. 
These parameters need to be declared and can be fixed to a value in this block.
All other fixed values for the model can be specified in the \$FIXED block.

```{r Parameters_and_Fixed}
parameters <- fixed[is.element(fixed_vars, ODE_vars)]
if(length(parameters)>0){
  print("$PARAM")
  parameters
}

fixed2 <- fixed[!is.element(fixed_vars, ODE_vars)]
fixed_vars2 <- fixed_vars[!is.element(fixed_vars, ODE_vars)]

all_RHS_vars <- c(ODE_vars, unique(unlist(RHS1)))
fixed3 <- fixed2[is.element(fixed_vars2, all_RHS_vars)]
if(length(fixed3)>0){
  print("$FIXED")
  fixed3
}
```

### Find \$TABLE variables
***mrgsolve block***  
mrgsolve has a \$TABLE block to define variables that are calculated after the 
ODEs advance.

*** Assumptions ***  
Any assignments that aren't already covered should be in the \$TABLE block.

```{r findTableParameters}
table <- c(table1, table2)
if(length(table)>0){
  print("$TABLE")
  table
}
```

### Find \$CAPTURE parameters
*** mrgsolve block ***  
mrgsolve uses the \$CAPTURE block to return values from the simulation. These 
are typically derived variables that combine other parameters and amounts.

*** Assumptions ***  
Looking for LHS variables that are not used on the RHS of 
assignments i.e. end of the computational chain.

*** Conversion details ***  

```{r findCapture}
## Look for lines where table variables are used on the RHS
pattern <- "\\(|\\)|\\*|\\/|\\-|\\+|\\>"
usedVars <- table %>%
  stringr::word(start=2, sep=" *= *") %>%
  stringr::word(start=1, sep=" *#|;") %>%
  stringr::str_replace_all(pattern = pattern, replacement = " ") %>%
  stringr::str_split(pattern = "[:space:]") %>%
  unlist() %>%
  unique() %>%
  .[.!=""]

defined_lines <- table %>% 
  stringr::word(1,sep=" *= *") %>%
  is.element(.,usedVars)

capture <- table[!defined_lines]
capture_vars <- stringr::word(capture,1,sep=" *= *")
if(length(capture)>0){
  print("$CAPTURE")
  capture_vars
}
```

## Print out
```{r print}
out <- c("$PROB","Lanzieri et al CMV vaccine strategy model")

if(length(parameters)>0){
  out <- c(out, "$PARAM",
           parameters %>%
             stringr::str_trim() 
  )
}

if(length(inits)>0){
  out <- c(out, "$INIT",
           inits %>%
             stringr::str_trim()
  )
}

if(length(cmts)>0){
  out <-  c(out, "$CMT",
            cmts %>%
              stringr::str_trim()
  )
}

if(length(fixed3)>0){
  out <-  c(out, "$FIXED",
            fixed3 %>% 
              stringr::str_trim()
  )
}

if(length(c(main1, main2))>0){
  out <- c(out, "$MAIN",
           c(main1, main2) %>%
             paste("double ", ., " ;", sep="") 
  )
}

if(length(ode)>0){
  out <-  c(out, "$ODE",
            ode %>%
              paste(., " ;", sep="") 
  )
}

if(length(table)>0){
  out <- c(out, "$TABLE",
           table %>%
             paste("double ", ., " ;", sep="")
  )
}

if(length(capture)>0){
  out <- c(out, "$CAPTURE",
           capture_vars %>%
             stringr::str_trim() 
  )
}

out
```

## Write to file
```{r write_to_file}
fileConn<-file("Calcium_Osc.cpp")
writeLines(out, fileConn)
close(fileConn)
```

## Test model file with mrgsolve
```{r test}
Sys.setenv(PATH=paste0("c:/program files/R/R-3.4.1/bin/x64/;",
                       "c:/RTools/bin/;",
                       "c:/RTools/mingw_64/bin/;",
                       Sys.getenv("PATH")))

library(mrgsolve)
mod <- mread("Calcium_Osc")
```